Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2018-01-04 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
final resubmission version (minor post knit edits in word)

When single insecticide use was investigated (Figs 4-5) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance, whereas higher values for cost led to slower evolution of resistance (Fig 5B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to Figure 2 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

@@

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? There are two main explanations for this apparent difference, a) dominance and b) polygenic resistance. The effect of dominance is that poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 4C, 8). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research on how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer. Secondly, if resistance is predominantly polygenic rather than monogenic poor implementation is likely to promote resistance by allowing the build up of many genes of small effect [@Gressel2017]. Indeed modelling work from 15 years ago [@Gardner1998][@Gardner1999] proposes an alternation of low and high doses with the low doses limiting the evolution of monogenic resistance and the high doses limiting that of the polygenic resistance. More research on the relative importance of monogenic and polygenic resistance for vector control would help here.

>>>>

When single insecticide use was investigated (Figs 4-5) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution of resistance, whereas higher values for cost led to slower evolution of resistance (Fig 5B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to Figure 2 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

@@

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? There are two main potential explanations for this apparent difference, a) dominance and b) polygenic resistance. The effect of dominance is that poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 4C, 8). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research on how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer. Secondly, if resistance is predominantly polygenic rather than monogenic poor implementation is likely to promote resistance by allowing the build up of many genes of small effect [@Gressel2017]. Indeed modelling work from 15 years ago [@Gardner1998][@Gardner1999] proposes an alternation of low and high doses with the low doses limiting the evolution of monogenic resistance and the high doses limiting that of the polygenic resistance. More research on the relative importance of monogenic and polygenic resistance for vector control would help here.
